COVID-19 and the 2020 U.S. Presidential Election: Counties with Extra COVID-19 Deaths Showed Less Support For Trump

### 2023-12-21 outputs/paper/paper.qmd

There are plenty of researches investigating the impacts of COVID-19 on 2020 Election with various perspectives. Baccini et al.(2021) suggested that Trump could win without COVID-19, but simply using the COVID-19 infection rate as the only factor contributing to his loss would be too naive. Socio-economic factors like high education attainment and race diversity are also essential in explaining the popular voting patterns. Besides, their research emphasized the significant impact of COVID-19 on those urban areas where there were no stay-at-home orders, particularly for the "swing" states. In addition, Noland and Zhang (2021) suggested that the deaths per case (DPC) rate is a more appropriate metric than the infection rate or mortality rate when analyzing the impact of COVID-19 on voting for Trump. However, they highlight the challenge in assessing COVID-19's impact on voting, as we may not know when the voters made their decision. In contrast, Clarke, Stewart, and Ho (2021) conducted surveys before and after the election, suggesting that COVID-19 is but not the dominant factor affecting the voting. The U.S.'s polarized political landscape might be more significant. Similarly, Hart (2021) also conducted a survey to ask about people's attitudes toward Trump. All the participants show natural or negative possession. They also find that some social movements like "Black Lives Matter" may have potential influence on the voting.

While all the above research investigated the correlation between COVID-19 and voting for Trump, previous studies predominantly employed models to identify variables significantly influencing Trump's vote share and perform counterfactual analyses to estimate his voting performance in the absence of COVID-19 or reduction in deaths. Instead, this study will employ the propensity score matching to find the causal effects between COVID-19 and Trump's lost. Furthermore, Wallace et al. (2023) highlights the the extra deaths during the pandemic are closely associated to a county's political preference. Building on the findings of Noland and Zhang (2021), the primary metric of COVID-19 in this analysis will be the excess death rate, specifically a high Death per Case (DPC) rate. However, given all counties have experienced COVID-19, this paper will first find the optimal treatment group(s). Then we will also conduct a counterfactual analysis to see if Trump could be re-elected with the treatment effect eliminated.

@@

@tbl-covid compares the COVID-19 impacts and income levels across the counties that each party won at 2020. Even though Republicans won in approximately five-sixths of the counties, counties supporting Biden reported nearly as ten times high as the average number of COVID-19 cases and deaths compared to those counties supporting Trump. Despite similar infection and mortality rates between the two groups, there is still a notable higher death-to-case rate in counties where Democrats won. These patterns suggest that the counties voting for Democrats are more severely impacted by COVID-19. Besides, Biden won the election with about 500 counties, this also indicate that these counties are densely populated urban areas and hence have more electoral votes. We can verify this from @fig-map, which compares the relative ratio of votes for the two parties and the death to cases rate in maps. It is clear that more counties had a preference for the Republicans, but the counties with higher DPC rates show a preference for Democrats. These counties, such as California and New York, are more affluent. In contrast, the states that are less impacted by COVID-19, such as Utah, vote more for Republicans.

@@

```{r}
#| fig-cap: The correlation between DPC rate with proportons of people with private insurance and without health insurance for the two parties
#| label: fig-insurance
#| fig-width: 8
#| fig-height: 4

@@

@fig-bar compares the ratio of the White population (race diversity) and education levels for the two parties. Interestingly, it suggests that counties with higher racial diversity are less likely to vote for Biden, as opposed to the trend observed for Republican-leaning counties, where increased diversity aligns with increased support. The average share of votes for Trump even reach 0.9 in the counties with the highest proportion of white residents. In terms of the ratio of people with high education attainment, since the average vote for Republicans is always higher than for Democrats, this indicate that educated people are more likely to vote for Trump. These observations are in line with the conclusions drawn by Baccini et al. (2021).

@@

Typically, the treatment should be defined before conducting the analysis, and this aims to reduce the risk of p-hacking [@Frost2023PHacking]. However, given the objective of this paper, which is to find whether extra COVID-19 deaths leads to Trump's loss in the last Election, pre-defining the treatment is challenging as all U.S. counties experienced COVID-19. Therefore, this study begins with exploratory analysis to identify the treatment group exhibiting the most significant differences in voting patterns for Trump. We will establish specific cutoffs for classifying counties as having "high" or "low" death per case rates (DPC). In addition, from @fig-dpc-income, low and high-income counties also seem to have different voting behaviours. Therefore, both DPC and income will be considered when choosing the optimal treatment.

Nevertheless, choosing cutoffs is critical, and we need to avoid p-hacking or data dredging risks. P-hacking means the manipulation of data analysis until we can find statistically significant results [@Frost2023PHacking]. If we simply choose one cutoff and see the significant results that match common sense, it may raise the p-hacking risk, and the analyses will be less convincing and comprehensive. Therefore, we will employ a grid of cutoffs for each variable, examining how treatment effect varies. For example, one treatment group can be counties with death per case and income levels higher than 0.3 quantiles. Using this way, we are not only able to find the optimal treatment group eventually, but avoid the risk of p-hacking.

@@

The propensity score is the probability an observation would have been treated based on the existing covariates using logistic regression. With propensity scores, it can help us to reduce the dimension [@Zhao2021PropensityR], meaning that we can describe an observation by a single value instead of multiple covariates. Another significant advantage of PSM is design separation [@Zhao2021PropensityR], meaning that PSM can separate the covariates balancing and effects estimation. This is especially beneficial as we can observe from @fig-insurance that the distribution of people with and without (private) insurance is unbalanced for the counties with a high and low DPC rate. Using propensity scores, we can match two observations with similar propensity scores, but from treatment and control groups, separately. We can then estimate the treatment effect based on many matched pairs with different propensity score. Therefore, propensity score mathcing can solve the limitation that we can not estimate causal effects from observational studies.

@@

@fig-dpc shows the correlation between votes for Trump and income levels, segmented into "high DPC" and "low DPC" counties at varying cutoffs. For instance, a cutoff of 0.3 classifies counties with a DPC rate below 0.3 quantile as "low DPC" and those above as "high DPC." By comparing these groups across different cutoffs, we aim to identify the cutoff where the difference in voting for Trump across different income levels is most pronounced. From the graph, we can find that, except for the cutoff 0.4, the two lines converge as income increases (right tail). Only for the 0.4 cutoff, the lines of the two groups converge at the middle income but do not overlap and diverge at the tails. Additionally, regardless of the cutoff applied, there is a common pattern that the support for Trump increases in poorer counties up to around the eighth bin, which is approximately the 0.4 income quantile, and then decreases in wealthier counties. This indicates a significant voting disparity between counties below and above the 0.4 income quantile.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: The Correlation Between Voting for Trump and Income Levels, Categorized by DPC Rate at Varied Quantile Cutoffs
#| label: fig-dpc
#| message: false
#| warning: false

@@

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: The Correlation Between Voting for Trump and DPC Rate, Categorized by Income Levels at Varied Quantile Cutoffs
#| label: fig-income
#| message: false
#| warning: false

@@

@fig-treatment shows a simple example when we test different cutoffs for DPC rate and income. From the graph, we can see that no matter which cutoff we set, the low DPC and high DPC counties always have a negative value of slope. The only difference is the absolute values of the slope. Conversely, the impact of income levels shows notable variability. For instance, at a cutoff of the 0.4 quantiles, low and high-income groups display opposite voting patterns for Trump. However, when we increase the cutoff to the 0.9 quantile, both income groups negatively correlate with Trump's vote share. This underscores the instability of income.

@@

# Base plot with cubic regression and vertical lines
p1 <- ggplot(binned_data, aes(x = dpc_pctile)) +
  geom_smooth(aes(y = mean_rep), colour = "black", method = "lm", 
              formula = y ~ poly(x, 3), se = FALSE, size = 1) +
  geom_vline(xintercept = 65, color = "lightgrey", linetype = "dashed", size = 0.8) +
  geom_vline(xintercept = 40, color = "#E69F00", linetype = "dashed", size = 0.8) +
  geom_smooth(data = left_data, aes(y = mean_rep, colour = "65"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) +
  geom_smooth(data = right_data, aes(y = mean_rep, colour = "65"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) + 
  geom_smooth(data = left_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  geom_smooth(data = right_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("65" = "lightgrey", "40" = "#E69F00"),
                       name = "Cutoffs",
                       labels = c("65" = "0.65 Quantile", "40" = "0.4 Quantile")) +
  theme(legend.position = "bottom",
          axis.title.y = element_blank()) +
  labs(x = "DPC Rate Percentiles", y = "Votes Share (%)")

@@

p2 <- ggplot(binned_data2, aes(x = income_pctile)) +
  geom_smooth(aes(y = mean_rep), colour = "black", 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE, size = 1) +
  geom_vline(xintercept = 90, color = "lightgrey", linetype = "dashed", size = 0.8) +
  geom_vline(xintercept = 40, color = "#E69F00", linetype = "dashed", size = 0.8) +
  geom_smooth(data = left_data, aes(y = mean_rep, colour = "90"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) +
  geom_smooth(data = right_data, aes(y = mean_rep, colour = "90"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) + 
  geom_smooth(data = left_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  geom_smooth(data = right_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("90" = "lightgrey", "40" = "#E69F00"),
                       name = "Cutoffs",
                       labels = c("90" = "0.90 Quantile", "40" = "0.4 Quantile")) +
  theme(legend.position = "bottom",
          axis.title.y = element_blank()) +
  labs(x = "Income Percentiles", y = "Votes Share (%)")

@@

To test the performance of PSM in balancing the covariates, @tbl-balance summarizes each variable's mean values in the control group and standard mean difference before and after the matching. Notably, there is a significant decrease in the standard mean difference for each variable, indicating that the PSM effectively reduced the imbalance between the treatment and control groups. Besides, we can see that the mean values in the control group approximately remain the same after the matching. This denote the reliability of the matching and suggest that PSM has successfully aligned the groups based on the observed covariates.

```{r}
#| label: tbl-balance
#| tbl-cap: The balance of each covariates before and after the mathcing
# Load necessary libraries
library(knitr)
library(kableExtra)

@@

@tbl-counter provides a comparative summary of the actual and counterfactual voting results in eleven swing states during the 2020 U.S. Presidential election. The counterfactual votes is calculated based on the treatment effects. That said, given other variables constant, Trump will have about 1.8% more votes in the counties with a DPC rate higher than 0.4 quantile, whereas Democrat will lose the same amount of votes in these counties. In reality, Trump only won three swing states and the total electoral votes was 232. However, the counterfactual analysis suggests that Trump would have won five additional states if the DPC rate disparities were eliminated, bringing him an extra 52 electoral votes. Then his total electoral votes will be 284, surpassing the required 270-vote threshold for reelection. Therefore, if Trump can eliminate the extra deaths, he could be re-elected.

@@

From @sec-data, it seems that a county's COVID death per case rate (DPC) and income levels are the two factors most related to voting for Trump. Therefore, the treatment will only incorporate these two factors and our objective is to define the cutoffs that maximize the difference between the treatment and control group. However, from @fig-dpc and @fig-income, it seems that the effect of income on the voting for Trump is highly dependent on the choice of cutoff as it appears a quadratic relationship. A quadratic pattern is not desired as different cutoffs will bring significant differences in the treatment effects. Furthermore, from @fig-treatment, we can observe that the two groups show an opposite relation to voting when the cutoff is low but the same relation for a higher cutoff. In contrast, DPC shows a general cubic relationship to Trump's vote; hence, the choice of cutoff will not generate huge different treatment effects. We can verify this from @fig-income, where no matter how the cutoff for defining "high" income changes, there will always be a linear or at least a closely linear relationship between them and the shift in cutoff will not bring a dramatic change in treatment effects (see @fig-treatment). Besides, from @fig-income, we can observe that the "high" and "low" DPC rate counties behave most differently with a cutoff at 0.4 quantiles. Therefore, the optimal treatment group is counties with a DPC rate higher than the 0.4 quantile.

To verify our conclusions, @tbl-income and @tbl-dpc compare the treatment effects for income and DPC at different cutoffs. From @tbl-income, we can observe that the treatment effects for income are unstable at different cutoffs. It increases gradually as we increase the cutoff and even shows a positive value if we set it to 0.9. This indicates that the voting difference for Trump between the high and low-income counties increases as cutoff increases. However, the reason why a cutoff 0.9 will generate a postive treatment effect is the both high and low income groups shows a negative correlation to the vote for Trump (see @fig-treatment). Hence we should not incorporate income into the treatment.

```{r}
#| label: tbl-income
#| tbl-cap: The treatment effect of income under different cutoffs (exlude income from calculation of propensity scores)
# show TE with different income levels
generate_rr_summaries <- function(data, cutoffs, treatment_var) {
  rr_list <- list()
  for (cutoff in cutoffs) {
    data_te <- data %>%
      mutate(high_dpc = ifelse(dpc > quantile(dpc, cutoff), 1, 0),
             high_income = ifelse(mean_household_income > quantile(mean_household_income, cutoff), 1, 0))
    data_te$treatment <- data_te[[treatment_var]]
    if (treatment_var == "high_income"){
      highinf_model <- glm(treatment ~ prop_higher_education +
                           no_insurance + private_insurance + males +
                           white_pct + black_pct, family = binomial(), data = data_te)
    }
    else{
      highinf_model <- glm(treatment ~ prop_higher_education + pctile +
                           no_insurance + private_insurance + males +
                           white_pct + black_pct, family = binomial(), data = data_te)
    }
    data_te$prop_score <- predict(highinf_model, type = "response")
    rr <- Match(Y = data_te$pct_vote_rep, Tr = data_te$treatment, X = data_te$prop_score, M = 1)
    rr_list[[as.character(cutoff)]] <- rr
  }
  summaries <- list()

@@

df_income <- extract_data_from_summaries(summaries_income)
df_dpc <- extract_data_from_summaries(summaries_dpc)
df_income %>% 
  rename(`Cutoffs` = cutoffs,
       `Treatment Effect` = estimate,
       `P value` = p,
       `Obs.` = on,
       `Treat. Obs.` = ot,
       `Treat. Obs` = mo,
       `Treat. Obs. Unw.` = mou) %>% 
kable(booktabs = TRUE) %>% 
add_header_above(c(" " = 3,
                   "Original" = 2, "Matched" = 2))
```

Compared to the income, the treatment effect for DPC at different cutoffs behaves more stable, where all values are negative and around -0.01. Besides, even though the number of treatment observations is the same as income, the number of unweighted treatment observations when setting treatment using DPC is much larger than income. This represents a higher number of samples, and therefore more accurate estimates. In addition, even though the optimal cutoff we found, 0.4 quantiles, does not have the maximum treatment effect, it is the only statistically significant one given 0.05 significance level.

```{r}
#| label: tbl-dpc
#| tbl-cap: The treatment effect of income under different cutoffs

df_dpc %>% 
  rename(`Cutoffs` = cutoffs,
       `Treatment Effect` = estimate,
       `P value` = p,
       `Obs.` = on,
       `Treat. Obs.` = ot,
       `Treat. Obs` = mo,
       `Treat. Obs. Unw.` = mou) %>% 
kable(booktabs = TRUE) %>% 
add_header_above(c(" " = 3,
                   "Original" = 2, "Matched" = 2))
```

@@

The candidates, especially the incumbent president, are always blamed or credited for the national economy and direction. But the 2020 Election was quite different because the voters needed to consider an extra variable "COVID-19" when making decisions. Most existing research indicates that Trump was criticized more than praised due to COVID-19, meaning that people only focus on COVID-19 and ignore the positive things that Trump made. In this paper, we see from @tbl-psm and @tbl-counter that the counties with deaths higher than 0.4 quantiles have significantly fewer votes for Trump, and Trump can win an extra five swing states if he can diminish the inequality. However, even though we did not incorporate income into the treatment group, we can not ignore its impacts on voting for Trump, as shown on @fig-dpc-income. The wealthy counties indeed vote less for Trump.

@@

```{r}
#| fig-cap: The comparision of Trump's voting behavior of counties with extra COVID-19 death in 2016 and 2020 Election
#| label: fig-compare
#| message: false
#| fig-width: 8
#| fig-height: 5

@@

On November 8, 2022, the 2022 United States Election, which is also the Midterm Election, was held. All seats of the U.S. House of Representatives and 35 seats of 100 were contested to determine the new Congress [@2022USElection]. Historically, people will have a relatively low support for the incumbent president, and hence the incumbent president's party will lose a notable number of seats. However, even though the Republican had overperformance in traditional Democrat states such as California and New York, the expected "Red Wave" was not appeared, with the Republican only taking nine more seats from Democrat in the Lower House [@2022USElection]. This is not be good news for Trump and his party, as in the 2018 Midterm Election, the expected "Blue Wave" appeared with Democrats taking 45 seats from the Republicans [@2018USElection]. Even though COVID was not officially ended and still had impacts on the U.S. in last year miterm election, it seems that the decreasing effect of COVID-19 did not bring an expected increase in support for Trump, especially the swing states, as we predicted in @tbl-counter.

>>>>

There are plenty of researchers investigating the impacts of COVID-19 on 2020 Election with various perspectives. Baccini et al.(2021) suggested that Trump could win without COVID-19, but simply using the COVID-19 infection rate as the only factor contributing to his loss would be too naive. Socio-economic factors like high education attainment and race diversity are also essential in explaining the popular voting patterns. Besides, their research emphasized the significant impact of COVID-19 on those urban areas where there were no stay-at-home orders, particularly for the "swing" states. In addition, Noland and Zhang (2021) suggested that the deaths per case (DPC) rate is a more appropriate metric than the infection rate or mortality rate when analyzing the impact of COVID-19 on voting for Trump. However, they highlight the challenge in assessing COVID-19's impact on voting, as we may not know when the voters made their decision. In contrast, Clarke, Stewart, and Ho (2021) conducted surveys before and after the election, suggesting that COVID-19 is but not the dominant factor affecting the voting. The U.S.'s polarized political landscape might be more significant. Similarly, Hart (2021) also conducted a survey to ask about people's attitudes toward Trump. All the participants show natural or negative possession. They also find that some social movements like "Black Lives Matter" may have potential influence on the voting.

While all the above research investigated the correlation between COVID-19 and voting for Trump, previous studies predominantly employed models to identify variables significantly influencing Trump's vote share and perform counterfactual analyses to estimate his voting performance in the absence of COVID-19 or reduction in deaths. Instead, this study will employ the propensity score matching to find the causal effects between COVID-19 and Trump's lost. Furthermore, Wallace et al. (2023) highlights the extra deaths during the pandemic are closely associated to a county's political preference. Building on the findings of Noland and Zhang (2021), the primary metric of COVID-19 in this analysis will be the excess death rate, specifically a high Death per Case (DPC) rate. However, given all counties have experienced COVID-19, this paper will first find the optimal treatment group(s). Then we will also conduct a counterfactual analysis to see if Trump could be re-elected with the treatment effect eliminated.

@@

@tbl-covid compares the COVID-19 impacts and income levels across the counties that each party won in 2020. Even though Republicans won in approximately five-sixths of the counties, counties supporting Biden reported nearly as ten times high as the average number of COVID-19 cases and deaths compared to those counties supporting Trump. Despite similar infection and mortality rates between the two groups, there is still a notable higher death-to-case rate in counties where Democrats won. These patterns suggest that the counties voting for Democrats are more severely impacted by COVID-19. Besides, Biden won the election with about 500 counties, this also indicate that these counties are densely populated urban areas and hence have more electoral votes. We can verify this from @fig-map, which compares the relative ratio of votes for the two parties and the death to cases rate in maps. It is clear that more counties had a preference for the Republicans, but the counties with higher DPC rates show a preference for Democrats. These counties, such as California and New York, are more affluent. In contrast, the states that are less impacted by COVID-19, such as Utah, vote more for Republicans.

@@

```{r}
#| fig-cap: The correlation between DPC rate with proportions of people with private insurance and without health insurance for the two parties
#| label: fig-insurance
#| fig-width: 8
#| fig-height: 4

@@

@fig-bar compares the ratio of the White population (race diversity) and education levels for the two parties. Interestingly, it suggests that counties with higher racial diversity are less likely to vote for Biden, as opposed to the trend observed for Republican-leaning counties, where increased diversity aligns with increased support. The average share of votes for Trump even reached 0.9 in the counties with the highest proportion of white residents. In terms of the ratio of people with high education attainment, since the average vote for Republicans is always higher than for Democrats, this indicate that educated people are more likely to vote for Trump. These observations are in line with the conclusions drawn by Baccini et al. (2021).

@@

Typically, the treatment should be defined before conducting the analysis, and this aims to reduce the risk of p-hacking [@Frost2023PHacking]. However, given the objective of this paper, which is to find whether extra COVID-19 deaths leads to Trump's loss in the last Election, pre-defining the treatment is challenging as all U.S. counties experienced COVID-19. Therefore, this study begins with exploratory analysis to identify the treatment group exhibiting the most significant differences in voting patterns for Trump. We will establish specific cut-offs for classifying counties as having "high" or "low" death per case rates (DPC). In addition, from @fig-dpc-income, low and high-income counties also seem to have different voting behaviours. Therefore, both DPC and income will be considered when choosing the optimal treatment.

Nevertheless, choosing cut-offs is critical, and we need to avoid p-hacking or data dredging risks. P-hacking means the manipulation of data analysis until we can find statistically significant results [@Frost2023PHacking]. If we simply choose one cutoff and see the significant results that match common sense, it may raise the p-hacking risk, and the analyses will be less convincing and comprehensive. Therefore, we will employ a grid of cut-offs for each variable, examining how treatment effect varies. For example, one treatment group can be counties with death per case and income levels higher than 0.3 quantiles. Using this way, we are not only able to find the optimal treatment group eventually, but avoid the risk of p-hacking.

@@

The propensity score is the probability an observation would have been treated based on the existing covariates using logistic regression. With propensity scores, it can help us to reduce the dimension [@Zhao2021PropensityR], meaning that we can describe an observation by a single value instead of multiple covariates. Another significant advantage of PSM is design separation [@Zhao2021PropensityR], meaning that PSM can separate the covariates balancing and effects estimation. This is especially beneficial as we can observe from @fig-insurance that the distribution of people with and without (private) insurance is unbalanced for the counties with a high and low DPC rate. Using propensity scores, we can match two observations with similar propensity scores, but from treatment and control groups, separately. We can then estimate the treatment effect based on many matched pairs with different propensity score. Therefore, propensity score matching can solve the limitation that we cannot estimate causal effects from observational studies.

@@

@fig-dpc shows the correlation between votes for Trump and income levels, segmented into "high DPC" and "low DPC" counties at varying cut-offs. For instance, a cutoff of 0.3 classifies counties with a DPC rate below 0.3 quantile as "low DPC" and those above as "high DPC." By comparing these groups across different cut-offs, we aim to identify the cutoff where the difference in voting for Trump across different income levels is most pronounced. From the graph, we can find that, except for the cutoff 0.4, the two lines converge as income increases (right tail). Only for the 0.4 cutoff, the lines of the two groups converge at the middle income but do not overlap and diverge at the tails. Additionally, regardless of the cutoff applied, there is a common pattern that the support for Trump increases in poorer counties up to around the eighth bin, which is approximately the 0.4 income quantile, and then decreases in wealthier counties. This indicates a significant voting disparity between counties below and above the 0.4 income quantile.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: The Correlation Between Voting for Trump and Income Levels, Categorized by DPC Rate at Varied Quantile Cut-offs
#| label: fig-dpc
#| message: false
#| warning: false

@@

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-cap: The Correlation Between Voting for Trump and DPC Rate, Categorized by Income Levels at Varied Quantile Cut-offs
#| label: fig-income
#| message: false
#| warning: false

@@

@fig-treatment shows a simple example when we test different cut-offs for DPC rate and income. From the graph, we can see that no matter which cutoff we set, the low DPC and high DPC counties always have a negative value of slope. The only difference is the absolute values of the slope. Conversely, the impact of income levels shows notable variability. For instance, at a cutoff of the 0.4 quantiles, low and high-income groups display opposite voting patterns for Trump. However, when we increase the cutoff to the 0.9 quantile, both income groups negatively correlate with Trump's vote share. This underscores the instability of income.

@@

# Base plot with cubic regression and vertical lines
p1 <- ggplot(binned_data, aes(x = dpc_pctile)) +
  geom_smooth(aes(y = mean_rep), colour = "black", method = "lm", 
              formula = y ~ poly(x, 3), se = FALSE, size = 1) +
  geom_vline(xintercept = 65, color = "lightgrey", linetype = "dashed", size = 0.8) +
  geom_vline(xintercept = 40, color = "#E69F00", linetype = "dashed", size = 0.8) +
  geom_smooth(data = left_data, aes(y = mean_rep, colour = "65"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) +
  geom_smooth(data = right_data, aes(y = mean_rep, colour = "65"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) + 
  geom_smooth(data = left_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  geom_smooth(data = right_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("65" = "lightgrey", "40" = "#E69F00"),
                       name = "Cut-offs",
                       labels = c("65" = "0.65 Quantile", "40" = "0.4 Quantile")) +
  theme(legend.position = "bottom",
          axis.title.y = element_blank()) +
  labs(x = "DPC Rate Percentiles", y = "Votes Share (%)")

@@

p2 <- ggplot(binned_data2, aes(x = income_pctile)) +
  geom_smooth(aes(y = mean_rep), colour = "black", 
              method = "lm", formula = y ~ poly(x, 3), se = FALSE, size = 1) +
  geom_vline(xintercept = 90, color = "lightgrey", linetype = "dashed", size = 0.8) +
  geom_vline(xintercept = 40, color = "#E69F00", linetype = "dashed", size = 0.8) +
  geom_smooth(data = left_data, aes(y = mean_rep, colour = "90"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) +
  geom_smooth(data = right_data, aes(y = mean_rep, colour = "90"), 
                     method = "lm", formula = y ~ x, se = FALSE, linetype = "dashed", size = 0.8) + 
  geom_smooth(data = left_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  geom_smooth(data = right_data_40, aes(y = mean_rep, colour = "40"), 
                method = "lm", formula = y ~ x, se = FALSE, linetype = "longdash", size = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("90" = "lightgrey", "40" = "#E69F00"),
                       name = "Cut-offs",
                       labels = c("90" = "0.90 Quantile", "40" = "0.4 Quantile")) +
  theme(legend.position = "bottom",
          axis.title.y = element_blank()) +
  labs(x = "Income Percentiles", y = "Votes Share (%)")

@@

To test the performance of PSM in balancing the covariates, @tbl-balance summarizes each variable's mean values in the control group and standard mean difference before and after the matching. Notably, there is a significant decrease in the standard mean difference for each variable, indicating that the PSM effectively reduced the imbalance between the treatment and control groups. Besides, we can see that the mean values in the control group approximately remain the same after the matching. This denotes the reliability of the matching and suggest that PSM has successfully aligned the groups based on the observed covariates.

```{r}
#| label: tbl-balance
#| tbl-cap: The balance of each covariate before and after the matching
# Load necessary libraries
library(knitr)
library(kableExtra)

@@

@tbl-counter provides a comparative summary of the actual and counterfactual voting results in eleven swing states during the 2020 U.S. Presidential election. The counterfactual votes is calculated based on the treatment effects. That said, given other variables constant, Trump will have about 1.8% more votes in the counties with a DPC rate higher than 0.4 quantile, whereas Democrats will lose the same number of votes in these counties. In reality, Trump only won three swing states and the total electoral votes was 232. However, the counterfactual analysis suggests that Trump would have won five additional states if the DPC rate disparities were eliminated, bringing him an extra 52 electoral votes. Then his total electoral votes will be 284, surpassing the required 270-vote threshold for re-election. Therefore, if Trump can eliminate the extra deaths, he could be re-elected.

@@

From @sec-data, it seems that a county's COVID death per case rate (DPC) and income levels are the two factors most related to voting for Trump. Therefore, the treatment will only incorporate these two factors and our objective is to define the cut-offs that maximize the difference between the treatment and control group. However, from @fig-dpc and @fig-income, it seems that the effect of income on the voting for Trump is highly dependent on the choice of cutoff as it appears a quadratic relationship. A quadratic pattern is not desired as different cut-offs will bring significant differences in the treatment effects. Furthermore, from @fig-treatment, we can observe that the two groups show an opposite relation to voting when the cutoff is low but the same relation for a higher cutoff. In contrast, DPC shows a general cubic relationship to Trump's vote; hence, the choice of cutoff will not generate huge different treatment effects. We can verify this from @fig-income, where no matter how the cutoff for defining "high" income changes, there will always be a linear or at least a closely linear relationship between them and the shift in cutoff will not bring a dramatic change in treatment effects (see @fig-treatment). Besides, from @fig-income, we can observe that the "high" and "low" DPC rate counties behave most differently with a cutoff at 0.4 quantiles. Therefore, the optimal treatment group is counties with a DPC rate higher than the 0.4 quantile.

To verify our conclusions, @tbl-income and @tbl-dpc compare the treatment effects for income and DPC at different cut-offs. From @tbl-income, we can observe that the treatment effects for income are unstable at different cut-offs. It increases gradually as we increase the cutoff and even shows a positive value if we set it to 0.9. This indicates that the voting difference for Trump between the high and low-income counties increases as cutoff increases. However, the reason why a cutoff 0.9 will generate a postive treatment effect is the both high- and low-income groups shows a negative correlation to the vote for Trump (see @fig-treatment). Hence we should not incorporate income into the treatment.

```{r}
#| label: tbl-income
#| tbl-cap: The treatment effect of income under different cut-offs (exlude income from calculation of propensity scores)
# show TE with different income levels
generate_rr_summaries <- function(data, cutoffs, treatment_var) {
  rr_list <- list()
  for (cutoff in cutoffs) {
    data_te <- data %>%
      mutate(high_dpc = ifelse(dpc > quantile(dpc, cutoff), 1, 0),
             high_income = ifelse(mean_household_income > quantile(mean_household_income, cutoff), 1, 0))
    data_te$treatment <- data_te[[treatment_var]]
    if (treatment_var == "high_income"){
      highinf_model <- glm(treatment ~ prop_higher_education +
                           no_insurance + private_insurance + males +
                           white_pct + black_pct, family = binomial(), data = data_te)
    }
    else{
      highinf_model <- glm(treatment ~ prop_higher_education + pctile +
                           no_insurance + private_insurance + males +
                           white_pct + black_pct, family = binomial(), data = data_te)
    }
    data_te$prop_score <- predict(highinf_model, type = "response")
    rr <- Match(Y = data_te$pct_vote_rep, Tr = data_te$treatment, X = data_te$prop_score, M = 1)
    rr_list[[as.character(cutoff)]] <- rr
  }
  summaries <- list()

@@

df_income <- extract_data_from_summaries(summaries_income)
df_dpc <- extract_data_from_summaries(summaries_dpc)
df_income %>% 
  rename(`Cut-offs` = cutoffs,
       `Treatment Effect` = estimate,
       `P value` = p,
       `Obs.` = on,
       `Treat. Obs.` = ot,
       `Treat. Obs` = mo,
       `Treat. Obs. Unw.` = mou) %>% 
kable(booktabs = TRUE) %>% 
add_header_above(c(" " = 3,
                   "Original" = 2, "Matched" = 2))
```

Compared to the income, the treatment effect for DPC at different cut-offs behaves more stable, where all values are negative and around -0.01. Besides, even though the number of treatment observations is the same as income, the number of unweighted treatment observations when setting treatment using DPC is much larger than income. This represents a higher number of samples, and therefore more accurate estimates. In addition, even though the optimal cutoff we found, 0.4 quantiles, does not have the maximum treatment effect, it is the only statistically significant one given 0.05 significance level.

```{r}
#| label: tbl-dpc
#| tbl-cap: The treatment effect of income under different cut-offs

df_dpc %>% 
  rename(`Cut-offs` = cutoffs,
       `Treatment Effect` = estimate,
       `P value` = p,
       `Obs.` = on,
       `Treat. Obs.` = ot,
       `Treat. Obs` = mo,
       `Treat. Obs. Unw.` = mou) %>% 
kable(booktabs = TRUE) %>% 
add_header_above(c(" " = 3,
                   "Original" = 2, "Matched" = 2))
```

@@

The candidates, especially the incumbent president, are always blamed or credited for the national economy and direction. But the 2020 Election was quite different because the voters needed to consider an extra variable "COVID-19" when making decisions. Most existing research indicates that Trump was criticized more than praised due to COVID-19, meaning that people only focus on COVID-19 and ignore the positive things that Trump made. In this paper, we see from @tbl-psm and @tbl-counter that the counties with deaths higher than 0.4 quantiles have significantly fewer votes for Trump, and Trump can win an extra five swing states if he can diminish the inequality. However, even though we did not incorporate income into the treatment group, we cannot ignore its impacts on voting for Trump, as shown on @fig-dpc-income. The wealthy counties indeed vote less for Trump.

@@

```{r}
#| fig-cap: The comparison of Trump's voting behavior of counties with extra COVID-19 death in 2016 and 2020 Election
#| label: fig-compare
#| message: false
#| fig-width: 8
#| fig-height: 5

@@

On November 8, 2022, the 2022 United States Election, which is also the Midterm Election, was held. All seats of the U.S. House of Representatives and 35 seats of 100 were contested to determine the new Congress [@2022USElection]. Historically, people will have a relatively low support for the incumbent president, and hence the incumbent president's party will lose a notable number of seats. However, even though the Republican had overperformance in traditional Democrat states such as California and New York, the expected "Red Wave" was not appeared, with the Republican only taking nine more seats from Democrat in the Lower House [@2022USElection]. This is not good news for Trump and his party, as in the 2018 Midterm Election, the expected "Blue Wave" appeared with Democrats taking 45 seats from the Republicans [@2018USElection]. Even though COVID was not officially ended and still had impacts on the U.S. in last year's mid-term election, it seems that the decreasing effect of COVID-19 did not bring an expected increase in support for Trump, especially the swing states, as we predicted in @tbl-counter.
